Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-10 talk/talk.tex
Work on talk.

\begin{frame}{the steady ice geometry problem (SIGP)}

@@

\vspace{-1mm}
\begin{itemize}
\item a glacier which is in balance with its climate
    \begin{itemize}
    \item the simplest such problem (grounded, nonsliding, isothermal)
    \item re~\emph{steady}: solve this and you can handle implicit steps too
    \end{itemize}
\item steady data given on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item determine: surface elevation $s(x,y)$ on $\Omega$, domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow, so there is an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
    $$\bu|_s \cdot \bn_s + a=0 \qquad \text{\emph{steady state}}$$

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \end{itemize}
\item \emph{off} the ice we know $a\le 0$ in steady state
\end{itemize}

\bigskip
\begin{columns}
    \column{0.6\textwidth}
        \begin{itemize}
        \item for all of this talk:
            \begin{itemize}
            \item $s$ is assumed to be well-defined (no overhangs)
            \item $s=b$ where ice is not present, so $s$ is defined on all of $\Omega$
            \end{itemize}
        \end{itemize}
    \column{0.4\textwidth}
        \includegraphics[width=1.05\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

\begin{frame}{SIGP strong form}

\begin{itemize}
\item the SIGP free-boundary problem determines $s$ on $\Omega$
\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}

@@

@@

\begin{itemize}
\item almost no one is solving the SIGP directly
    \begin{itemize}
    \item or doing implicit time-stepping (\emph{basically the same as SIGP})
    \item except Wirbel \& Jarosch \cite{WirbelJarosch2020} (\emph{semi-coupled, nonscalable, artificial ice layer})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item mostly with shallow approximations, e.g.~PISM run below
    \end{itemize}
\end{itemize}

\medskip
\begin{columns}
    \column{0.8\textwidth}
        \hfill \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.2\textwidth}
        \scriptsize \emph{show clip of Alpine run by Julien Seguinot}
\end{columns}
\end{frame}

@@

\begin{block}{definition: ice dynamics operator}
$\Phi : s \mapsto - \bus \cdot \bn_s$

@@

\hfill \includegraphics[width=0.75\textwidth]{figs/idoaction.png}
\end{block}

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding $F_{\Lambda_s}(\bu,p) = 0$:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz \cite{JouvetRappaz2011})
    \item optimal solver exists (Isaac et al \cite{IsaacStadlerGhattas2015})
    \end{itemize}
\end{itemize}

@@

\begin{frame}{SIGP weak form is a variational inequality}

\begin{itemize}
\item using $\Phi$ gives a cleaner NCP over $\Omega$:
\begin{align*}
s - b &\ge 0 \\
\Phi(s) - a &\ge 0 \\
(s - b) (\Phi(s) - a) &= 0
\end{align*}

@@

\begin{block}{SIGP weak form}
find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
    $$\hspace{10mm} F(s)[r - s] \ge \ip{a}{r-s}$$
for all $r \in \mathcal{K}$, where $F(s)[w] = \int_\Omega \Phi(s)\, w \,dx dy$
\end{block}

\item well-posedness is completely open
    \begin{itemize}
    \item but existence holds for SIA version (Jouvet \& Bueler \cite{JouvetBueler2012})
    \end{itemize}

\end{itemize}

\begin{frame}{goal: robust and optimal SIGP solver}

\begin{block}{numerical goal}
Given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$.  Use the numerical solution of the Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$.  Solve VI for $s^h \in \mathcal{K}^h$:

$$F(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
\end{block}

\begin{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$: work is $O(m)$
        \begin{itemize}
        \item[{\color{black} $\circ$}] optimal requires multigrid
        \end{itemize}
    \item<2> \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{what thing is $\Phi$?}

\begin{block}{SIGP weak form}
find $s \in \mathcal{K}$ so $F(s)[r - s] \ge \ip{a}{r-s}$ for all $r \in \mathcal{K}$
\end{block}

\begin{itemize}
\item $F(s)[w] = \int_\Omega \Phi(s)\, w$ is dual of $\Phi(s)=-\bus\cdot\bn_s$
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item some kind of integral operator (\emph{convolve with bdry Stokeslets})
    \item $\Phi$ is concentrated near the diagonal, i.e.~short range?
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
\small
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$
\normalsize
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FIXME \cite{BuelerMitchell2022}
\end{itemize}

\includegraphics[width=0.8\textwidth]{figs/extruded.png}

@@

@@

\begin{frame}{References}
%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

@@

@@

\bibitem{JouvetBueler2012}
G.~Jouvet and E.~Bueler. \textbf{Steady, shallow ice sheets as obstacle problems: well-posedness and finite element approximation.} {\em SIAM J. Appl. Math.}, 72(4):1292--1314, 2012.

\bibitem{JouvetRappaz2011}
G.~Jouvet and J.~Rappaz. \textbf{Analysis and finite element approximation of a nonlinear stationary Stokes problem arising in glaciology.} {\em Adv. Numer. Anal.}, 2011.

>>>>

\begin{frame}{the ice geometry problem (IGP)}

@@

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
\item take simplest such problem (grounded, nonsliding, isothermal)
\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow
\item $\therefore$ must solve an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a & &\text{\emph{evolving}} \\
\bu|_s \cdot \bn_s + a &= 0 & &\text{\emph{steady state}}
\end{align*}

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \item \emph{not} on the ice: $a\le 0$ for $t>0$ (or in steady state)
    \end{itemize}
\end{itemize}

\bigskip
\begin{columns}
    \column{0.55\textwidth}
        \begin{itemize}
        \item assumed for this talk:
            \begin{itemize}
            \item $s$ well-defined (no overhangs)
            \item $s=b$ where ice is not present
            \item $s$ is defined on all of $\Omega$
            \end{itemize}
        \end{itemize}
    \column{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{figs/simpledomain.png}
\end{columns}
\end{frame}

\begin{frame}{steady IGP strong form}

\begin{itemize}
\item the steady IGP free-boundary problem for $s(x,y)$ on $\Omega$
\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}

@@

\begin{frame}{\alert{implicit} IGP strong form}

\begin{itemize}
\item the evolving IGP free-boundary problem for $s(x,y,t)$ on $\Omega \times [0,T]$
\item consider one step of backward Euler scheme
    \begin{itemize}
    \item solve for $s,\bu,p$ at new time $t^\ell$
    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$, and write $s$ for $s^\ell$
    \end{itemize}
\item again: NCP for $s$ coupled to a Stokes problem for $\bu,p$:

\begin{align*}
s - b &\ge 0 && \text{on $\Omega$} \\
\alert{s - s^{\ell-1} - \Delta t\, \bu|_s \cdot \bn_s - \Delta t\, a} &\ge 0 && \text{''} \\
(s - b) (\alert{s - s^{\ell-1} - \Delta t\, \bu|_s \cdot \bn_s - \Delta t\, a}) &= 0 && \text{''} \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\bu &= \bzero && \text{on $\Gamma_0$} \\
\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$}
\end{align*}

    \begin{itemize}
    \item \alert{pretty similar \dots mostly suffices to consider steady IGP}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Brinkerhoff et al (2017) (\emph{upwinding limits time-step?})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
    \item neither scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

\medskip
\begin{columns}
    \column{0.7\textwidth}
        \includegraphics[width=0.8\textwidth]{figs/seguinot.png}
    \column{0.3\textwidth}
        \scriptsize \emph{by Julien Seguinot}
\end{columns}

\medskip
    \item my goal is to do better
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{block}{ice dynamics operator}
$\Phi : s \mapsto - \bus \cdot \bn_s$

@@

\hfill \includegraphics[width=0.7\textwidth]{figs/idoaction.png}
\end{block}

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
\end{itemize}

@@

\begin{frame}{IGP weak form is a variational inequality}

\begin{itemize}
\item using $\Phi$ gives a cleaner NCP over $\Omega$ for steady IGP:
\begin{align*}
s - b &\ge 0 \\
\Phi(s) - a &\ge 0 \\
(s - b) (\Phi(s) - a) &= 0
\end{align*}

@@

\begin{block}{steady IGP weak form}
find admissible $s \in \mathcal{K} = \{r \ge b\} \subset W^{1,\qq}(\Omega)$ so that
    $$\ip{\Phi(s)-a}{r-s} \ge 0$$
for all $r \in \mathcal{K}$
\end{block}

\item well-posedness is completely open
    \begin{itemize}
    \item but existence holds for SIA version (Jouvet \& Bueler, 2012)
    \end{itemize}

\end{itemize}
\end{frame}

\begin{frame}{what kind of beast is $\Phi$?}

\begin{center}
\mbox{\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}}
\end{center}

\begin{itemize}
\small
\item $\Phi : \mathcal{K} \subset W^{1,\qq}(\Omega) \to W^{-1,\pp}(\Omega)$ where $\qq=4$ and $\pp=4/3$
\item the Stokes stress balance is nonlocal so $\Phi$ is ``dense''
    \begin{itemize}
    \item $\Phi$ is some kind of integral operator? (\emph{convolve with Stokeslets})
    \item $\Phi$ is short range? (\emph{concentrated near the diagonal})
    \end{itemize}
\item $\Phi \approx \Phi_{\text{SIA}}$ for shallow glaciers, a differential operator:
    $$\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right]$$

    \begin{itemize}
    \item $\Phi$ is diffusive in the large!
    \item $\Phi$ is degenerate at the ice margin!
    \end{itemize}
\item yet common to say $\Phi$ is hyperbolic because SKE is transport equation for thickness: $H_t + \overline{\bu}\, \grad H = a$
\end{itemize}
\end{frame}

\begin{frame}{goal: robust and optimal IGP solver}

\begin{block}{numerical goal}
given triangulation $\mathcal{T}$ of $\Omega$ and $P_1$ space $\mathcal{V}^h \subset W^{1,\qq}(\Omega)$, and convex set $\mathcal{K}^h = \{r^h \ge b^h\} \subset \mathcal{V}^h$, solve VI for $s^h \in \mathcal{K}^h$:

$$\ip{\Phi^h(s^h) - a}{r^h - s^h} \ge 0 \qquad \text{for all $r^h \in \mathcal{K}^h$}$$
\end{block}

\begin{itemize}
\item[]
    \begin{itemize}
    \item $P_1$ because low regularity at free boundary
    \item uses numerical solution of Stokes problem $F_{\Lambda_{s^h}}(\bu^h,p^h)=0$ to evaluate $\Phi^h(s^h)=-\bu|_{s^h}\cdot\bn_{s^h}$
    \end{itemize}
\item seeking a solver which is:
    \begin{itemize}
    \item \alert{robust}
    \item \alert{optimal} over $m$ degrees of freedom in $\mathcal{V}^h$ 
        \begin{itemize}
        \item[{\color{black} $\circ$}] optimal means work is $O(m)$ \dots requires multigrid
        \end{itemize}
    \item<2> \emph{for now these are aspirations \dots}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Stokes on an extruded mesh}

\begin{itemize}
\item FIXME
\end{itemize}

\includegraphics[width=0.8\textwidth]{figs/extruded.png}
\end{frame}

@@

\begin{itemize}
\item FIXME
\end{itemize}

\includegraphics[width=0.5\textwidth]{figs/decompclassical.png}

@@

\begin{frame}{smoothers}

\begin{itemize}
\item FIXME
\end{itemize}
\end{frame}

@@

\begin{frame}{Some references}
%\bibliography{../paper/msg.bib}
%\bibliographystyle{plain}

@@

\hrulefill

\bibitem{Brinkerhoffetal2017}
D.~Brinkerhoff, M.~Truffer, and A.~Aschwanden. \textbf{Sediment transport drives tidewater glacier periodicity.} {\em Nature Communications}, 8(1):1--8, 2017.

@@
